### 2019-04-25 introduction.tex
second revision fixes up to data manipulation

Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland and is serviced by Isavia~\cite{Isavia_about}. The airport currently has two functioning runways operating in both directions and are designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West). 
There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over eight million passengers and 62.300 air transport movements in 2017~\cite{isavia_facts_2017}, where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}. For 2018 the passenger numbers show a yearly growth of 13,44\% (at the time of writing) for the same time period from the previous year~\cite{isavia_pass_statistics_2018}. The increase in traffic at Keflavík Airport can and does experience constrained capacity of the runways during peak traffic periods, here airport capacity is defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology}.

A significant constraint of airport capacity is caused by distance separation regulations between aircraft in flight due to wake turbulence. The term "wake turbulence" describes the effect of the rotating air masses generated behind the wing tips of large jet aircraft~\cite{doc4444full}. A smaller aircraft can experience loss of control in close proximity behind a larger aircraft because of those swirling air masses. The general flight safety requirements for maintaining safe distance between aircraft are established by the International Civil Aviation Organisation (ICAO)~\cite{doc4444full}. The European Aviation Safety Agency (EASA) provides oversight of those requirements and their application by the Air Traffic Management (ATM).

The distance separation due to wake turbulence is especially critical during take-off and on approach for landing. ICAO has specified wake turbulence separation minima requirements that are applied to aircraft in the flight-path of another aircraft. Those requirements classify each aircraft in a Wake Turbulence Category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft. Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in Nautical Miles (NM). Generally the ICAO wake categories are three - Heavy, Medium and Light. The classification is based primarily on the weight of the aircraft, but this can vary slightly for different countries.

The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are some 50 years old. Even though those separations are considered reliable, they are often conservative and outdated in view of newer aircraft models and as such generate unnecessary over-separation in some cases.
In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an intergovernmental organisation assisting in safe and efficient air traffic operations.

@@

Apart from wake turbulence separation, the other major factor that affects capacity is Runway Occupancy Time (ROT), which is the time interval an aircraft occupies a runway during landing or departure. The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, per se, does limit the throughput of a runway. The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport. The peak traffic hours are determined by the time intervals when the airport operates at high capacity loads.

@@

The aim of this study is to estimate the constraints for implementing the RECAT-EU wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when is the runway occupancy a limiting factor for increased capacity.

>>>>

Keflavík International Airport (IATA: KEF, ICAO: BIKF) is the major gateway to Iceland and is operated by Isavia~\cite{Isavia_about}, a shareholding company solely owned by the Icelandic state. The airport currently has two functioning runways operating in both directions, designated as RWY-01 (South-North), RWY-19 (North-South), RWY-10 (West-East) and RWY-28 (East-West). 
There has been an unforeseen increase of passengers travelling through Keflavík Airport in recent years, with over 9,8~million passengers and 97.432 air transport movements in 2018~\cite{isavia_facts_2017}. Where a movement is defined as an aircraft take-off or landing at an airport~\cite{aircraft_movement}. For 2018 the passenger numbers show a yearly growth of 12.0\% from the previous year~\cite{isavia_pass_statistics_2018}. The increase in traffic at Keflavík Airport does experience constrained capacity of the runways during peak traffic periods, here airport capacity is defined as the number of movements per unit of time that can be accepted~\cite{airport_capacity_methodology} without causing delays on the ground or in the air.

A significant constraint of airport capacity is caused by distance separation regulations between aircraft in flight due to wake turbulence. The term "wake turbulence" describes the effect of the rotating air masses generated behind the wins of large jet aircraft~\cite{doc4444full}. A smaller aircraft can experience loss of control in close proximity behind a larger aircraft because of those swirling air masses. The general flight safety requirements for maintaining safe distance between aircraft are established by the International Civil Aviation Organisation (ICAO)~\cite{doc4444full}. The European Aviation Safety Agency (EASA) provides the specific definition and oversight of those requirements and their application by Air Traffic Management (ATM) providers in the European Economic Area (EAA). 
The distance separation due to wake turbulence is especially critical during take-off and on approach for landing. ICAO has specified wake turbulence separation minima requirements that are applied to aircraft in the flight-path of another aircraft. Those requirements classify each aircraft into a Wake Turbulence Category (WTC) and the wake turbulence separation is applied to a following aircraft associated with the wake category of a leading aircraft. Thus the aircraft form a pair composed of a "leader" and a "follower", that observe a certain spacing in Nautical Miles (NM). Generally the ICAO wake categories are three - Heavy, Medium and Light. The classification is based primarily on the weight of the aircraft, but this can vary slightly for different countries or regions such as the EEA.

The ICAO WTC categorisation is based on observations and calculations performed on aircraft models that are in some cases 50 years old. Even though those separations are considered reliable, they are often conservative and outdated in view of newer aircraft models and as such generate unnecessary over-separation in some cases.
In recent years, a wake turbulence re-categorisation scheme known as RECAT-EU~\cite{rooseleer2015recat} has been devised by European Organisation for the Safety of Air Navigation (EUROCONTROL)~\cite{EUROCONTROL_recat_eu}, an inter-European organisation with wide responsibility for ensuring safe and efficient air traffic operations in Europe including the operation of the Central Flow Management Unit (CFMU) in Brussels and the Maastricht Area Control Center.

@@

Apart from wake turbulence separation, the other major factor that affects capacity is Runway Occupancy Time (ROT), which is the time interval that an aircraft occupies the runway during landing or departure. The rule of thumb is that no aircraft can use a runway before the previous aircraft has vacated it and runway occupancy, consequently does limit the throughput of a runway. The emphasis in this project will be on runway occupancy times for arriving aircraft (AROT) during peak traffic hours at the airport. The peak traffic hours are determined by the time intervals when the airport operates at high capacity loads.

@@

The aim of this study is to estimate the constraints for implementing the RECAT-EU wake turbulence categorisation and separation minima for Keflavík International Airport, the effect this can produce for airfield throughput and when the runway occupancy is a limiting factor for increased capacity.
